---
title: 'Weekly Exercises #3'
author: "Anthony Palma"
output: 
  html_document:
    keep_md: TRUE
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_download: true
---


```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE, error=TRUE, message=FALSE, warning=FALSE)
```

```{r libraries}
library(tidyverse)     # for graphing and data cleaning
library(gardenR)       # for Lisa's garden data
library(lubridate)     # for date manipulation
library(ggthemes)      # for even more plotting themes
library(geofacet)      # for special faceting with US map layout
theme_set(theme_minimal())       # My favorite ggplot() theme :)
```

```{r data}
# Lisa's garden data
data("garden_harvest")

# Seeds/plants (and other garden supply) costs
data("garden_spending")

# Planting dates and locations
data("garden_planting")

# Tidy Tuesday data
kids <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-15/kids.csv')
```

## Warm-up exercises with garden data

These exercises will reiterate what you learned in the "Expanding the data wrangling toolkit" tutorial. If you haven't gone through the tutorial yet, you should do that first.

  1. Summarize the `garden_harvest` data to find the total harvest weight in pounds for each vegetable and day of week (HINT: use the `wday()` function from `lubridate`). Display the results so that the vegetables are rows but the days of the week are columns.

```{r}
garden_harvest %>% 
  mutate(day_wk = wday(date, label = TRUE)) %>% 
  group_by(vegetable, day_wk) %>% 
  summarize(total_weight = sum(weight * 0.00220462)) %>% 
  pivot_wider(id_cols = vegetable, 
               names_from = day_wk, 
              values_from = total_weight)

```

  2. Summarize the `garden_harvest` data to find the total harvest in pound for each vegetable variety and then try adding the plot from the `garden_planting` table. This will not turn out perfectly. What is the problem? How might you fix it?
  
Because I did it all at once joining by vegetable and variety, there is a row for each plot and now there are multiple vegetable total weights repeated because of wrong plots.  I could filter out repeats by starting with the garden_planting data first.  Then, I could group by vegetable and variety followed by summarizing the min date of harvest hoping to reduce those extra rows and reducing repeats.       

```{r}
garden_harvest %>%
    group_by(vegetable, variety) %>% 
    summarize(total_weight = sum(weight * 0.00220462)) %>%
    left_join(garden_planting, by = c("vegetable", "variety"))
```

  3. I would like to understand how much money I "saved" by gardening, for each vegetable type. Describe how I could use the `garden_harvest` and `garden_spending` datasets, along with data from somewhere like [this](https://products.wholefoodsmarket.com/search?sort=relevance&store=10542) to answer this question. You can answer this in words, referencing various join functions. You don't need R code but could provide some if it's helpful.
  

First, I would group_by my garden_harvest by vegetable and variety to get it down to vegetable type specifically separately grouped.  Then, I would left join the garden_spending data set to the garden_harvest set by vegetable and variety.  Next, I could make a small set of data with the whole foods pricing that has the different vegetable and type with its unit price next to it.  I would then again left_join this data set to the already joined data set prior by vegetable type and variety.  Lastly, I would perform some type of transformation that mutated creating another variable that multiplies the actual harvest weight by unit price to figure out how much we would have to spend at whole foods to purchase these. And finally, I would mutate and create another variable called cost difference that does (price with tax from garden spending) - (price of total weight of each veg variety at store after the multiplcation).  Any positive numbers are where we saved, negative are where we lost. 

I could also sum these columns to get one number portraying how much we lost or saved overall.  
  
  

  4. Subset the data to tomatoes. Reorder the tomato varieties from smallest to largest first harvest date. Create a barplot of total harvest in pounds for each variety, in the new order.

```{r}
garden_harvest %>% 
  filter(vegetable == "tomatoes") %>% 
  mutate(variety2 = fct_reorder(variety, date, first, .desc = FALSE)) %>% 
  group_by(variety2) %>% 
  summarize(count = n(),
            tot_weight_lbs = sum(weight * 0.00220462))   %>%
  arrange(variety2) %>%
ggplot() +
  geom_col(aes( y = variety2, x = tot_weight_lbs), fill = "blue") +
  labs(title = "Tot Harvest Wt of Vairiety of Tomatoes, in ascending order of harvest", x = "total wt, lbs", y = "variety of tomatoes")


```

  5. In the `garden_harvest` data, create two new variables: one that makes the varieties lowercase and another that finds the length of the variety name. Arrange the data by vegetable and length of variety name (smallest to largest), with one row for each vegetable variety. HINT: use `str_to_lower()`, `str_length()`, and `distinct()`.
  
```{r}
garden_harvest %>%
  mutate(variety_low = str_to_lower(variety)) %>%
  mutate(variety_lngth = str_length(variety)) %>%
  distinct(variety_lngth, .keep_all = TRUE) %>%
  arrange(vegetable, variety_lngth)
```

  6. In the `garden_harvest` data, find all distinct vegetable varieties that have "er" or "ar" in their name. HINT: `str_detect()` with an "or" statement (use the | for "or") and `distinct()`.

```{r}
library("stringr")

garden_harvest %>%
  group_by(vegetable, variety) %>%
  mutate(variety = str_to_lower(variety)) %>%
  distinct(variety, .keep_all = TRUE) %>%
  mutate(has_erar = str_detect(variety, "er | ar") )
```


## Bicycle-Use Patterns

In this activity, you'll examine some factors that may influence the use of bicycles in a bike-renting program.  The data come from Washington, DC and cover the last quarter of 2014.


Two data tables are available:

- `Trips` contains records of individual rentals
- `Stations` gives the locations of the bike rental stations

Here is the code to read in the data. We do this a little differently than usualy, which is why it is included here rather than at the top of this file. To avoid repeatedly re-reading the files, start the data import chunk with `{r cache = TRUE}` rather than the usual `{r}`.

```{r cache=TRUE}
data_site <- 
  "https://www.macalester.edu/~dshuman1/data/112/2014-Q4-Trips-History-Data.rds" 
Trips <- readRDS(gzcon(url(data_site)))
Stations<-read_csv("http://www.macalester.edu/~dshuman1/data/112/DC-Stations.csv")
```

**NOTE:** The `Trips` data table is a random subset of 10,000 trips from the full quarterly data. Start with this small data table to develop your analysis commands. **When you have this working well, you should access the full data set of more than 600,000 events by removing `-Small` from the name of the `data_site`.**

### Temporal patterns

It's natural to expect that bikes are rented more at some times of day, some days of the week, some months of the year than others. The variable `sdate` gives the time (including the date) that the rental started. Make the following plots and interpret them:

  7. A density plot, which is a smoothed out histogram, of the events versus `sdate`. Use `geom_density()`.
  
```{r}
Trips %>%
  ggplot() +
  geom_density(aes(sdate)) +
  labs(title = "Distriubtion of when bikes were rented by date", x = "date")
```


This tells me bike rentals skyrocket in the fall peaking in early to late October followed by a steady decline as it gets colder (w some intermittent spikes).  Then, as the holiday season approaches there is a slight uptick followed by a continual decline into the colder months.  

  
  8. A density plot of the events versus time of day.  You can use `mutate()` with `lubridate`'s  `hour()` and `minute()` functions to extract the hour of the day and minute within the hour from `sdate`. Hint: A minute is 1/60 of an hour, so create a variable where 3:30 is 3.5 and 3:45 is 3.75.
  
```{r}
Trips %>%
  mutate(time = (lubridate::hour(sdate) + lubridate::minute(sdate)/60)) %>%
  ggplot() +
  geom_density(aes(time)) +
  labs(title = "Distriubtion of when bikes were rented by time of day", x = "time")


#Trips %>%
  #mutate(timeofday = format(sdate, format = "%H:%M:%S")) %>%
  #mutate(time = round(time, digits = ))
```
  
  9. A bar graph of the events versus day of the week. Put day on the y-axis.
  
```{r}
Trips %>%
  mutate(day_wk = wday(sdate, label = TRUE)) %>%
  ggplot() +
  geom_bar(aes(day_wk), fill = "blue") +
  labs(title = "Distriubtion of when bikes were rented by day of wk", x = "date")
```
  
  10. Facet your graph from exercise 8. by day of the week. Is there a pattern?
  
  Weekdays appear to peak two times: morning around 9 /10 and then afternoon around 3/4.  Weekends appear to peak once w steady declines, peaking around noon each weekend day.
  
```{r}
Trips %>%
  mutate(time = (lubridate::hour(sdate) + lubridate::minute(sdate)/60)) %>%
  mutate(day_wk = wday(sdate, label = TRUE)) %>%
  ggplot() +
  geom_density(aes(time)) +
  facet_wrap(~day_wk) +
  labs(title = "Distriubtion of when bikes were rented by time of day faceted by day of wk", x = "time")

```
  
The variable `client` describes whether the renter is a regular user (level `Registered`) or has not joined the bike-rental organization (`Causal`). The next set of exercises investigate whether these two different categories of users show different rental behavior and how `client` interacts with the patterns you found in the previous exercises. 

  11. Change the graph from exercise 10 to set the `fill` aesthetic for `geom_density()` to the `client` variable. You should also set `alpha = .5` for transparency and `color=NA` to suppress the outline of the density function.
  
```{r}
Trips %>%
  mutate(time = (lubridate::hour(sdate) + lubridate::minute(sdate)/60)) %>%
  mutate(day_wk = wday(sdate, label = TRUE)) %>%
  ggplot() +
  geom_density(aes(time, fill = client), alpha = 0.5, color = NA) +
  facet_wrap(~day_wk) +
  labs(title = "Distriubtion of when bikes were rented by time of day faceted by day of wk", x = "time")
```

  12. Change the previous graph by adding the argument `position = position_stack()` to `geom_density()`. In your opinion, is this better or worse in terms of telling a story? What are the advantages/disadvantages of each?
  
  
  I think this is WORSE.  It takes away the general pattern of the larger to fit the smaller maybe?  All I know is this shows how one fits the other just to look at comparable amounts of density in certain areas which is an adv from the last that simply have the density distributions.  But a big disadvantage is that this take away from the gerneral trends of some of the clients on certain days.  Above, we can see both trends not attempting to fit the other.  Thus, this graph below may cause you to draw false conclusions about cetrain times because of the misleading stacking.  
  
```{r}
Trips %>%
  mutate(time = (lubridate::hour(sdate) + lubridate::minute(sdate)/60)) %>%
  mutate(day_wk = wday(sdate, label = TRUE)) %>%
  ggplot() +
  geom_density(aes(time, fill = client), alpha = 0.5, color = NA, position = position_stack()) +
  facet_wrap(~day_wk) +
  labs(title = "Distriubtion of when bikes were rented by time of day faceted by day of wk", x = "time")
```
  
  13. In this graph, go back to using the regular density plot (without `position = position_stack()`). Add a new variable to the dataset called `weekend` which will be "weekend" if the day is Saturday or Sunday and  "weekday" otherwise (HINT: use the `ifelse()` function and the `wday()` function from `lubridate`). Then, update the graph from the previous problem by faceting on the new `weekend` variable. 
  
```{r}
Trips %>%
  mutate(time = (lubridate::hour(sdate) + lubridate::minute(sdate)/60)) %>% 
  mutate(day_wk = wday(sdate, label = TRUE)) %>%
  mutate(weekend = ifelse(day_wk == "Sat | Sun", "weekend", "weekday")) %>%
  ggplot() +
  geom_density(aes(time, fill = client), alpha = 0.5, color = NA) +
  facet_wrap(~weekend) +
  labs(title = "Distriubtion of when bikes were rented by time of day faceted by weekend var", x = "time")
```
  
  14. Change the graph from the previous problem to facet on `client` and fill with `weekday`. What information does this graph tell you that the previous didn't? Is one graph better than the other?
  
```{r}
Trips %>%
  mutate(time = (lubridate::hour(sdate) + lubridate::minute(sdate)/60)) %>% 
  mutate(day_wk = wday(sdate, label = TRUE)) %>%
  mutate(weekend = ifelse(day_wk == "Sat | Sun", "weekend", "weekday")) %>%
  ggplot() +
  geom_density(aes(time, fill = weekend), alpha = 0.5, color = NA) +
  facet_wrap(~client) +
  labs(title = "Distriubtion of when bikes were rented by time of day faceted by weekend var", x = "time")
```
  
### Spatial patterns

  15. Use the latitude and longitude variables in `Stations` to make a visualization of the total number of departures from each station in the `Trips` data. Use either color or size to show the variation in number of departures. We will improve this plot next week when we learn about maps!
  
```{r}
Stations <- Stations %>%
  mutate(sstation = name)
  
```
  
```{r}
Trips %>%
  group_by(sstation) %>%
  left_join(Stations, by = "sstation") %>%
  ggplot() +
  geom_point(aes(lat, long, color = nbBikes)) +
  labs(title = "Variation in number of departures by long, lat of station location")
```
  
  
  16. Only 14.4% of the trips in our data are carried out by casual users. Create a plot that shows which area(s) have stations with a much higher percentage of departures by casual users. What patterns do you notice? (Again, we'll improve this next week when we learn about maps).
  
```{r}
Trips %>%
  group_by(sstation) %>%
  left_join(Stations, by = "sstation") %>%
  ggplot() +
  geom_point(aes(lat, long, color = nbBikes)) +
  facet_wrap(~client) +
  labs(title = "Variation in number of departures by long, lat of station location faceted by client")
```
  
### Spatiotemporal patterns

  17. Make a table with the ten station-date combinations (e.g., 14th & V St., 2014-10-14) with the highest number of departures, sorted from most departures to fewest. Save this to a new dataset and print out the dataset. Hint: `as_date(sdate)` converts `sdate` from date-time format to date format. 
  
```{r}

```
  
  18. Use a join operation to make a table with only those trips whose departures match those top ten station-date combinations from the previous part.
  
```{r}

```
  
  19. Build on the code from the previous problem (ie. copy that code below and then %>% into the next step.) and group the trips by client type and day of the week (use the name, not the number). Find the proportion of trips by day within each client type (ie. the proportions for all 7 days within each client type add up to 1). Display your results so day of week is a column and there is a column for each client type. Interpret your results.

**DID YOU REMEMBER TO GO BACK AND CHANGE THIS SET OF EXERCISES TO THE LARGER DATASET? IF NOT, DO THAT NOW.**

## GitHub link

  20. Below, provide a link to your GitHub page with this set of Weekly Exercises. Specifically, if the name of the file is 03_exercises.Rmd, provide a link to the 03_exercises.md file, which is the one that will be most readable on GitHub.

## Challenge problem! 

This problem uses the data from the Tidy Tuesday competition this week, `kids`. If you need to refresh your memory on the data, read about it [here](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-09-15/readme.md). 

  21. In this exercise, you are going to try to replicate the graph below, created by Georgios Karamanis. I'm sure you can find the exact code on GitHub somewhere, but **DON'T DO THAT!** You will only be graded for putting an effort into this problem. So, give it a try and see how far you can get without doing too much googling. HINT: use `facet_geo()`. The graphic won't load below since it came from a location on my computer. So, you'll have to reference the original html on the moodle page to see it.
  


**DID YOU REMEMBER TO UNCOMMENT THE OPTIONS AT THE TOP?**
