---
title: 'Weekly Exercises #4'
author: "Anthony Palma"
output: 
  html_document:
    keep_md: TRUE
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_download: true
---


```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE, error=TRUE, message=FALSE, warning=FALSE)
```

```{r libraries}
library(tidyverse)     # for data cleaning and plotting
library(lubridate)     # for date manipulation
library(openintro)     # for the abbr2state() function
library(palmerpenguins)# for Palmer penguin data
library(maps)          # for map data
library(ggmap)         # for mapping points on maps
library(gplots)        # for col2hex() function
library(RColorBrewer)  # for color palettes
library(sf)            # for working with spatial data
library(leaflet)       # for highly customizable mapping
library(carData)       # for Minneapolis police stops data
library(ggthemes)      # for more themes (including theme_map())
theme_set(theme_minimal())
```

```{r data}
# Starbucks locations
Starbucks <- read_csv("https://www.macalester.edu/~ajohns24/Data/Starbucks.csv")

starbucks_us_by_state <- Starbucks %>% 
  filter(Country == "US") %>% 
  count(`State/Province`) %>% 
  mutate(state_name = str_to_lower(abbr2state(`State/Province`))) 

# Lisa's favorite St. Paul places - example for you to create your own data
favorite_stp_by_lisa <- tibble(
  place = c("Home", "Macalester College", "Adams Spanish Immersion", 
            "Spirit Gymnastics", "Bama & Bapa", "Now Bikes",
            "Dance Spectrum", "Pizza Luce", "Brunson's"),
  long = c(-93.1405743, -93.1712321, -93.1451796, 
           -93.1650563, -93.1542883, -93.1696608, 
           -93.1393172, -93.1524256, -93.0753863),
  lat = c(44.950576, 44.9378965, 44.9237914,
          44.9654609, 44.9295072, 44.9436813, 
          44.9399922, 44.9468848, 44.9700727)
  )

#COVID-19 data from the New York Times
covid19 <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")

```


### Starbucks locations (`ggmap`)

  1. Add the `Starbucks` locations to a world map. Add an aesthetic to the world map that sets the color of the points according to the ownership type. What, if anything, can you deduce from this visualization?  
  
```{r}
world <- get_stamenmap(
    bbox = c(left = -180, bottom = -57, right = 179, top = 82.1), 
    maptype = "terrain",
    zoom = 2)  
ggmap(world) + 
  geom_point(data = Starbucks, 
             aes(x = Longitude, y = Latitude, color = `Ownership Type`), 
             alpha = .3, 
             size = .1) +
  theme_map() +
  theme(legend.background = element_blank())

```

  

  2. Construct a new map of Starbucks locations in the Twin Cities metro area (approximately the 5 county metro area). 
  
  
```{r}
twin_cities <- get_stamenmap(
    bbox = c(left = -93.9938, bottom = 44.6099, right = -92.3651, top = 45.2584), 
    maptype = "terrain",
    zoom = 10)
ggmap(twin_cities) +
  geom_point(data = Starbucks, 
             aes(x = Longitude, y = Latitude), 
             alpha = .5, 
             size = 2,
             color = "red") +
  theme_map() 
```
  
  
  

  3. In the Twin Cities plot, play with the zoom number. What does it do?  (just describe what it does - don't actually include more than one map).  
  
  
  
  It determines the level of detail displayed with higher zoom showing greater detail.  
  
  

  4. Try a couple different map types (see `get_stamenmap()` in help and look at `maptype`). Include a map with one of the other map types.  
  
```{r}
twin_cities_2 <- get_stamenmap(
    bbox = c(left = -93.9938, bottom = 44.6099, right = -92.3651, top = 45.2584), 
    maptype = "toner",
    zoom = 10)
twin_cities_3 <- get_stamenmap(
    bbox = c(left = -93.9938, bottom = 44.6099, right = -92.3651, top = 45.2584), 
    maptype = "watercolor",
    zoom = 10)
ggmap(twin_cities_2) +
  geom_point(data = Starbucks, 
             aes(x = Longitude, y = Latitude), 
             alpha = .5, 
             size = 2,
             color = "red") +
  theme_map()

ggmap(twin_cities_3) +
  geom_point(data = Starbucks, 
             aes(x = Longitude, y = Latitude), 
             alpha = .5, 
             size = 2,
             color = "red") +
  theme_map()
```

  
  
  
  

  5. Add a point to the map that indicates Macalester College and label it appropriately. There are many ways you can do think, but I think it's easiest with the `annotate()` function (see `ggplot2` cheatsheet).
  
  
```{r}
ggmap(twin_cities) +
  annotate(geom = "text", x = -93.169044, y = 44.937893, label = "Macalester College", color = "orange", size = 6) +
  annotate(geom = "point", x = -93.169044, y = 44.937893, size = 3, color = "navy") +
  theme_map()
```
  
  
  

### Choropleth maps with Starbucks data (`geom_map()`)

The example I showed in the tutorial did not account for population of each state in the map. In the code below, a new variable is created, `starbucks_per_10000`, that gives the number of Starbucks per 10,000 people. It is in the `starbucks_with_2018_pop_est` dataset.

```{r}
census_pop_est_2018 <- read_csv("https://www.dropbox.com/s/6txwv3b4ng7pepe/us_census_2018_state_pop_est.csv?dl=1") %>% 
  separate(state, into = c("dot","state"), extra = "merge") %>% 
  select(-dot) %>% 
  mutate(state = str_to_lower(state))

starbucks_with_2018_pop_est <-
  starbucks_us_by_state %>% 
  left_join(census_pop_est_2018,
            by = c("state_name" = "state")) %>% 
  mutate(starbucks_per_10000 = (n/est_pop_2018)*10000)
```

  6. **`dplyr` review**: Look through the code above and describe what each line of code does.
  
162 - reads in csv and saving data as `census_pop_est_2018`
163 -  separates the leading dot from the state name, and makes new column `dot` that is empty 
164 -  selects all variables except `dot` which was empty anyway
165 -  creates new variable `state` with all lower-case letters fromvthe previous state column 

167 -  creates new dataset name `starbucks_with_2018_pop_est` for changes to ocme
168 -  starts with `starbucks_us_by_state` dataset which was previously defined already
169 - left joins with `census_pop_est_2018` dataset
170 -  this is left joined directly throguh the common column of `state`
171 -  creates new variable to measure Starbucks locations per 10,000 population

  7. Create a choropleth map that shows the number of Starbucks per 10,000 people on a map of the US. Use a new fill color, add points for all Starbucks in the US (except Hawaii and Alaska), add an informative title for the plot, and include a caption that says who created the plot (you!). Make a conclusion about what you observe.
  
  
```{r}
states_map <- map_data("state")
starbucks_with_2018_pop_est %>% 
  ggplot() +
  geom_map(map = states_map,
           aes(map_id = state_name,
               fill = starbucks_per_10000)) +
  geom_point(data = Starbucks %>%
               filter(`Country` == "US") %>%
               filter(`State/Province` != "AK") %>% 
               filter(`State/Province` != "HI"),
             aes(x = Longitude, y = Latitude),
             size = .05,
             alpha = .2, 
             color = "yellow") +
  expand_limits(x = states_map$long, y = states_map$lat) + 
  labs(title = "Starbucks Locations in the United States by state",
       fill = "Locations per 10k People",
       caption = "Created by Anthony Palma") +
  scale_fill_viridis_c(option = "inferno") +
  theme_map() +
  theme(legend.background = element_blank()) 
```


### A few of your favorite things (`leaflet`)

  8. In this exercise, you are going to create a single map of some of your favorite places! The end result will be one map that satisfies the criteria below. 
  

```{r}
favorite_ap <- tibble(
  place = c("Macalester College", "Carowinds", "Home", "Lake Norman High School", 
            "Lake Norman", "Amstar 14",
            "The Whitehouse", "The Capitol", "Hometown", "Nicholson Field"),
  longitude = c(-93.169044,  -80.95387268066406, -80.934710,-80.9016627, -80.932986, -80.85735321044922,  -76.9818115234375, -77.00588989257812, -77.615214, -93.16988302491517),
  latitude = c(44.937893, 35.10505294799805, 35.568530, 35.5977068, 35.5923267, 35.58753967285156, 38.879150390625, 38.87559509277344, 43.157285, 44.93534610120574),
  top3 = c("no", "no","yes", "no", "no", "no", "yes", "no", "no", "yes"))
pal <- colorFactor(c("blue", "red"),
                     domain = c("no", "yes"))
favorite_ap
leaflet(data = favorite_ap) %>% 
  addProviderTiles(providers$Stamen.Terrain) %>% 
  addCircles(label = ~place,
             weight = 10,
             opacity = 1,
             color = ~pal(top3)) %>% 
  addPolylines(lng = ~longitude, 
               lat = ~latitude,
               color = col2hex("goldenrod"))
```

  
## Revisiting old datasets

This section will revisit some datasets we have used previously and bring in a mapping component. 

### Bicycle-Use Patterns

The data come from Washington, DC and cover the last quarter of 2014.

Two data tables are available:

- `Trips` contains records of individual rentals
- `Stations` gives the locations of the bike rental stations

Here is the code to read in the data. We do this a little differently than usualy, which is why it is included here rather than at the top of this file. To avoid repeatedly re-reading the files, start the data import chunk with `{r cache = TRUE}` rather than the usual `{r}`. This code reads in the large dataset right away.

```{r cache=TRUE}
data_site <- 
  "https://www.macalester.edu/~dshuman1/data/112/2014-Q4-Trips-History-Data.rds" 
Trips <- readRDS(gzcon(url(data_site)))
Stations<-read_csv("http://www.macalester.edu/~dshuman1/data/112/DC-Stations.csv")
```

  9. Use the latitude and longitude variables in `Stations` to make a visualization of the total number of departures from each station in the `Trips` data. Use either color or size to show the variation in number of departures. This time, plot the points on top of a map. Use any of the mapping tools you'd like.
  
```{r}
trips_new <- Trips %>% 
  rename(name = sstation) %>% 
  left_join(Stations, by = "name") %>% 
  group_by(name, lat, long) %>% 
  summarize(number_departures = n())

dc_map <- get_stamenmap(
    bbox = c(left = -77.2119, bottom = 38.7817, right = -76.8377, top = 39.0021), 
    maptype = "terrain",
    zoom = 12)
ggmap(dc_map) + 
  geom_point(data = trips_new, 
             aes(x = long, y = lat, color = number_departures), 
             alpha = .5) +
  theme_map() +
  labs(title = "DC Bike Rentals by Station",
       size = "Number of Departures")
```
  
  10. Only 14.4% of the trips in our data are carried out by casual users. Create a plot that shows which area(s) have stations with a much higher percentage of departures by casual users. What patterns do you notice? Also plot this on top of a map. I think it will be more clear what the patterns are.
  
```{r}
trips_casual <- 
  Trips %>%
  mutate(name = sstation) %>%
  left_join(Stations, by = c("name")) %>%
  group_by(lat, long, client) %>%
  summarize(client_depart_count = n()) %>%
  mutate(client_depart_prop = client_depart_count / sum(client_depart_count)) 

ggmap(dc_map) + 
  geom_point(data = trips_casual, 
             aes(x = long, y = lat, color = client_depart_prop), 
             alpha = .5) +
  theme_map() +
  labs(title = "Percentage of Casual Riders by Station",
       size = "Number of Departures")
```

Higher percentage of casual users are pretty evenly spread throughout DC.  Higher percentage locations appear to be near more atraction type areas for tourism.  

It appears casual users are more near tourist locations inner city whereas registered users who use more often appear to be more outside the city.  
  
  
  
### COVID-19 data

The following exercises will use the COVID-19 data from the NYT.

  11. Create a map that colors the states by the most recent cumulative number of COVID-19 cases (remember, these data report cumulative numbers so you don't need to compute that). Describe what you see. What is the problem with this map?
  
```{r}
covid_new <- covid19 %>% 
  group_by(state) %>% 
  mutate(state = str_to_lower(state)) %>% 
  summarize(total_cases = max(cases)) 
  
covid_new %>%  
  ggplot() +
  geom_map(map = states_map,
           aes(map_id = state,
               fill = total_cases)) +
  expand_limits(x = states_map$long, y = states_map$lat) + 
  theme_map() + 
  labs(title = "Covid-19 Case Cumulative Count by State", fill = "Cumulative Cases")
```
  
Because it is just raw counts, bigger states like CA, TX, FL, and NY will have higher counts.  This can correct that by showing percentage of population infected. 


  12. Now add the population of each state to the dataset and color the states by most recent cumulative cases/10,000 people. See the code for doing this with the Starbucks data. You will need to make some modifications. 
  
  
```{r}
covid_total_per_10k <- covid_new %>% 
  left_join(census_pop_est_2018,
            by = "state") %>% 
  mutate(cases_per_10k = (total_cases/est_pop_2018)*10000) 

covid_total_per_10k %>% 
  ggplot() +
  geom_map(map = states_map,
           aes(map_id = state,
               fill = cases_per_10k)) +
  expand_limits(x = states_map$long, y = states_map$lat) + 
  theme_map() + 
  labs(title = "Covid-19 Cases by State", fill = "Cumulative Cases per 10k")
```
  
  
  13. **CHALLENGE** Choose 4 dates spread over the time period of the data and create the same map as in exercise 12 for each of the dates. Display the four graphs together using faceting. What do you notice?
  
```{r}
covid_by_date <- covid19 %>% 
  filter(date %in% c(mdy("03/01/2020"), mdy("05/01/2020"), mdy("07/01/2020"), mdy("09/01/2020"))) %>% 
  group_by(state) %>% 
  mutate(state = str_to_lower(state)) %>% 
  left_join(census_pop_est_2018,
            by = "state") %>% 
  mutate(cases_per_10k = (cases/est_pop_2018)*10000) 

covid_by_date %>% 
  ggplot() +
  geom_map(map = states_map,
           aes(map_id = state,
               fill = cases_per_10k)) +
  expand_limits(x = states_map$long, y = states_map$lat) + 
  theme_map() + 
  facet_wrap(~ date) +
  theme(legend.position = "left") +
  labs(title = "Covid Cases by State over specific dates", fill = "Cumulative Cases per 10k People")
```
  
It is super cool how on 3/1 you can see where cases are and are not.  You can see as each 2 month window passes the state colors start to lighten indicating increasing cases specifically higher in the S/SE.  

## Minneapolis police stops

These exercises use the datasets `MplsStops` and `MplsDemo` from the `carData` library. Search for them in Help to find out more information.

  14. Use the `MplsStops` dataset to find out how many stops there were for each neighborhood and the proportion of stops that were for a suspicious vehicle or person. Sort the results from most to least number of stops. Save this as a dataset called `mpls_suspicious` and display the table.  
  
  
```{r}
mpls_suspicious <- 
  MplsStops %>% 
  group_by(neighborhood) %>% 
  summarize(num_stops = cumsum(n()), 
            prop_sus = mean(problem == "suspicious")) %>% 
  arrange(desc(num_stops))

mpls_suspicious
```
  
  
  15. Use a `leaflet` map and the `MplsStops` dataset to display each of the stops on a map as a small point. Color the points differently depending on whether they were for suspicious vehicle/person or a traffic stop (the `problem` variable). HINTS: use `addCircleMarkers`, set `stroke = FAlSE`, use `colorFactor()` to create a palette.  
  
```{r}
pal_15 <- colorFactor(c("red", "green"),
                    domain = c("suspicious", "traffic"))
MplsStops %>% 
  leaflet() %>% 
  addProviderTiles(providers$Stamen.Terrain) %>% 
  addCircles(lng = ~long, 
             lat = ~lat, 
             weight = .5,
             opacity = 1,
             color = ~pal_15(problem),
             stroke = FALSE)
```
  
  
  
  
  16. Save the folder from moodle called Minneapolis_Neighborhoods into your project/repository folder for this assignment. Make sure the folder is called Minneapolis_Neighborhoods. Use the code below to read in the data and make sure to **delete the `eval=FALSE`**. Although it looks like it only links to the .sph file, you need the entire folder of files to create the `mpls_nbhd` data set. These data contain information about the geometries of the Minneapolis neighborhoods. Using the `mpls_nbhd` dataset as the base file, join the `mpls_suspicious` and `MplsDemo` datasets to it by neighborhood (careful, they are named different things in the different files). Call this new dataset `mpls_all`.

```{r}
mpls_nbhd <- st_read("Minneapolis_Neighborhoods/Minneapolis_Neighborhoods.shp", quiet = TRUE)
```

```{r}
mpls_dem <- mpls_suspicious %>% 
  left_join(MplsDemo, by = "neighborhood")

mpls_all <- mpls_nbhd %>% 
  left_join(mpls_dem, by = c("BDNAME" = "neighborhood")) %>% 
  rename("neighborhood" = "BDNAME")
```




  17. Use `leaflet` to create a map from the `mpls_all` data  that colors the neighborhoods by `prop_suspicious`. Display the neighborhood name as you scroll over it. Describe what you observe in the map.
  


  
  18. Use `leaflet` to create a map of your own choosing. Come up with a question you want to try to answer and use the map to help answer that question. Describe what your map shows. 
  
  
  
  
## GitHub link

  19. Below, provide a link to your GitHub page with this set of Weekly Exercises. Specifically, if the name of the file is 04_exercises.Rmd, provide a link to the 04_exercises.md file, which is the one that will be most readable on GitHub.
  
  


**DID YOU REMEMBER TO UNCOMMENT THE OPTIONS AT THE TOP?**
